---
title: "Brazil Cropland"
author: "Nick McManus"
date: '2022-07-28'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sf)  # polygons
library(raster) #raster package
library(terra)  # newer/faster/better raster package
library(rnaturalearth)  #administrative boundaries data
library(exactextractr)  # extracting what's underneath polys, points
library(kableExtra)
library(tidyverse)
library(caret) # for confusion matrix
```

## ROI

Let's start with looking at Brazil, our region of interest

```{r}
# get brazil from rnaturalearth data
brazil <- ne_countries(
  scale = "medium",
  country = "Brazil",
  returnclass = "sf") %>% 
  dplyr::select(sovereignt)

# plot to check
ggplot() +
  geom_sf(data = brazil)
```

## Projected Cropland

### Chen et al. 2022
First read in projected 2020 cropland from Chen et al. 2022. Then, crop and mask this global projection to just Brazil. Finally, reclassify the raster to keep only pixels representing cropland.

Source: Chen, G., Li, X., & Liu, X. (2022). Global land projection based on plant functional types with a 1-km resolution under socio-climatic scenarios. Scientific Data, 9(1), 125. https://doi.org/10.1038/s41597-022-01208-6

```{r}
# read in GeoTiff for SSP1 scenario (other scenarios will be examined later)
chen_ssp1_2020_global <- rast("data/chen/global_PFT_SSP1_RCP26_2020.tif")

# transform brazil to crop chen raster bc it's faster than reprojecting the whole global raster
brazil_trans <- brazil %>% 
  st_transform(crs = crs(chen_ssp1_2020_global)) %>% 
  vect() # vect is more agreeable to rasts

# crop and mask chen data to brazil_trans vector, then project back to our desired crs
chen_ssp1_2020_brazil <- chen_ssp1_2020_global %>% 
  crop(brazil_trans) %>% 
  mask(brazil_trans) %>% 
  terra::project(y = crs(brazil))

# isolate just cropland from the raster
chen_crop_ssp1_2020_brazil <- chen_ssp1_2020_brazil
chen_crop_ssp1_2020_brazil[chen_crop_ssp1_2020_brazil != 18] <- 0 # if value isn't 18, make 0
chen_crop_ssp1_2020_brazil[chen_crop_ssp1_2020_brazil == 18] <- 10 # if value is 18, make 10

plot(chen_crop_ssp1_2020_brazil)
```
 
 
### Cao et al., 2021
This is where I will read in Cao data

Source: 
```{r}
cao_ssp1_2020_global <- rast("data/cao/globalCropland_2020CE_SSP1_RCP26.tif")


```







## Observed Cropland

For observed cropland in Brazil, we will use data from MapBiomas (Collection 6), which uses Landsat mosaics to classify LULC at 30m resolution. (More information and GeoTiff source found at: https://mapbiomas.org/en/colecoes-mapbiomas-1?cama_set_language=en)
Cropland data for analysis will be defined as "Agriculture" from the MapBiomas classification, which includes temporary and perennial crops but omits pasture and mosaic agriculture.

```{r}
# WARNING: below analysis (aggregate step) will take a long time to run!
# Once run, the last line of code in this chunk will save the intermediate raster locally.
# If this has been already run once, skip running this chunk and move on to next


# read in GeoTiff
mapbiomas_2020 <- rast("data/mapbiomas_brasil_coverage_2020.tif")

# change resolution from 30m to roughly 1km
# to figure out factor that gets res closest to target raster (Chen), use this:
    factor <- ceiling(res(chen_crop_ssp1_2020_brazil)[1]/res(mapbiomas_2020)[1])
    factor

# now change resolution using terra:aggregate()
mapbiomas_2020_agg <- aggregate(mapbiomas_2020,   #raster 
                            fact = 34,            #agg factor (positive int)
                            fun = "modal",        #categorical values in raster
                            na.rm = TRUE)         #don't drop entire aggregation bc of one NA

# resample lower res mapbiomas using Chen data to ensure same resolution (fine-tuning the res)
mapbiomas_2020_resamp <- resample(mapbiomas_2020_agg, chen_crop_ssp1_2020_brazil, method = "near")

# because crs of mapbiomas and chen rasters are different, repeat process of
# transforming brazil for purposes of cropping and masking
brazil_trans2 <- brazil %>% 
  st_transform(crs = crs(mapbiomas_2020_resamp)) %>% 
  vect() 

# crop and mask mapbiomas layer to ROI
mapbiomas_2020_cropped <- mapbiomas_2020_resamp %>% 
  crop(brazil_trans2, mask = TRUE)

# create reclassification matrix to combine and keep only cropland categories
reclass_df <- c(3, 0,   #forest formation
                4, 0,   #savanna formation
                5, 0,   #mangrove
                49, 0,  #wooded restinga
                11, 0,  #wetlands
                12, 0,  #grassland
                32, 0,  #salt flat
                29, 0,  #rocky outcrop
                13, 0,  #other nonforest formations
                15, 0,  #pasture
                9, 0,   #forest plantation
                21, 0,  #mosaic ag and pasture
                23, 0,  #beach, dune, and sand spot
                24, 0,  #urban area
                30, 0,  #mining
                25, 0,  #other non vegetated areas
                33, 0,  # river, lake, ocean
                31, 0,  #aquaculture
                27, 0,  #non observed
                39, 1,  #soybean
                20, 1,  #sugar cane
                40, 1,  #rice
                41, 1,  #other temp crops
                46, 1,  #coffee
                47, 1,  #citrus
                48, 1)  #other perennial crops

# reshape df into matrix with columns and rows
reclass_m <- matrix(reclass_df, ncol = 2, byrow = TRUE)

# use terra:classify() to reclassify raster values based on matrix
mapbiomas_2020_reclass <- classify(mapbiomas_2020_cropped, reclass_m)

# use if you want to assign all 0 pixels to no data value:
#mapbiomas_reclass[mapbiomas_reclass == 0] <- NA

# save intermediate raster to save on computing time in future
writeRaster(mapbiomas_2020_reclass, "data/mapbiomas_brasil_2020_cropland.tif")

```


```{r}
# skip steps above (after run once) and read in saved raster below to save time
mapbiomas_crop_reclass <- rast("data/mapbiomas_brasil_2020_cropland.tif") 

# crs do not match, so use terra::project() to match with chen raster
mapbiomas_crop_2020 <- project(mapbiomas_crop_reclass, crs(chen_crop_ssp1_2020_brazil), method = "near")

# checking that crs of mapbiomas raster matches with brazil (desired crs)
plot(mapbiomas_crop_2020)
plot(brazil, col = NA, add = TRUE)
```



## Confusion matrix

Now we will compare projects vs observed cropland within Brazil. 

### Chen vs. Mapbiomas
First, the Chen and MapBiomas rasters will be added to determine overlap between crop classifications. 
```{r}
# add chen and mapbiomas rasters 
chen_mapbiomas_crop_2020 <- mapbiomas_crop_2020 + chen_crop_ssp1_2020_brazil
plot(chen_mapbiomas_crop_2020)

# return frequency of each value (0, 1, 10, 11) in new raster
counts <- freq(chen_mapbiomas_crop_2020)
counts

# break up freq values ba
nocrop <- counts[1, 3]
crop_mapbiomas_only <- counts[2, 3]
crop_chen_only <- counts[3,3]
crop_both <- counts[4,3]

# determine total number of pixels with values for calculating percentages later
total <- nocrop + crop_mapbiomas_only + crop_chen_only + crop_both
total



# break down into values for confusion matrix

# number of real positive cases (P), pixels for which mapbiomas observed cropland
p <- crop_mapbiomas_only + crop_both
# number of real negative cases (N), pixels for which mapbiomas did not observe cropland
n <- nocrop + crop_chen_only

# true positive
tp <- crop_both
# false positive
fp <- crop_chen_only
# true negative
tn <- nocrop
# false negative
fn <- crop_mapbiomas_only

#percentages:
# sensitivity (true positive rate)
tpr <- tp/p
# specificity (true negative rate)
tnr <- tn/n
# false positive rate
fpr <- fp/n
# false negative rate
fnr <- fn/p


conf_matrix <- matrix(c(tp, fn, fp, tn), ncol=2, byrow=TRUE)
colnames(conf_matrix) <- c("Predicted Positive", "Predicted Negative")
rownames(conf_matrix) <- c("Observed Positive", "Observed Negative")
conf_matrix

conf_matrix_perc <- matrix(c(tpr, fnr, fpr, tnr), ncol=2, byrow=TRUE)
colnames(conf_matrix_perc) <- c("Predicted Positive", "Predicted Negative")
rownames(conf_matrix_perc) <- c("Observed Positive", "Observed Negative")
conf_matrix_perc

```


## Checking baseline

Checking the sensitivity and specificity of Chen vs Mapbiomas for 2015, both considered "observed" models.

First read in, crop, mask, and reclassify 2015 Chen raster
```{r}
# read in GeoTiff for 2015 Chen
chen_2015_global <- rast("data/chen/global_PFT_2015.tif")


# crop and mask chen data to brazil_trans vector, then project back to our desired crs
chen_2015_brazil <- chen_2015_global %>% 
  crop(brazil_trans) %>% 
  mask(brazil_trans) %>% 
  terra::project(y = crs(brazil))

# isolate just cropland from the raster
chen_crop_2015_brazil <- chen_2015_brazil
chen_crop_2015_brazil[chen_crop_2015_brazil != 18] <- 0 # if value isn't 18, make 0
chen_crop_2015_brazil[chen_crop_2015_brazil == 18] <- 10 # if value is 18, make 10

#plot(chen_crop_2015_brazil)
```

Next read in, crop, mask, and reclassify 2015 MapBiomas raster
```{r}
# WARNING: below analysis (aggregate step) will take a long time to run!
# Once run, the last line of code in this chunk will save the intermediate raster locally.
# If this has been already run once, skip running this chunk and move on to next


# read in GeoTiff
mapbiomas_2015 <- rast("data/mapbiomas_brasil_coverage_2015.tif")

# change resolution from 30m to roughly 1km
# to figure out factor that gets res closest to target raster (Chen), use this:
    factor <- ceiling(res(chen_crop_2015_brazil)[1]/res(mapbiomas_2015)[1])
    factor

# now change resolution using terra:aggregate()
mapbiomas_2015_agg <- aggregate(mapbiomas_2015,   #raster 
                            fact = 34,            #agg factor (positive int)
                            fun = "modal",        #categorical values in raster
                            na.rm = TRUE)         #don't drop entire aggregation bc of one NA

# resample lower res mapbiomas using Chen data to ensure same resolution (fine-tuning the res)
mapbiomas_2015_resamp <- resample(mapbiomas_2015_agg, chen_crop_2015_brazil, method = "near")

# crop and mask mapbiomas layer to ROI
mapbiomas_2015_cropped <- mapbiomas_2015_resamp %>% 
  crop(brazil_trans2, mask = TRUE)

# use terra:classify() to reclassify raster values based on matrix
# use same matrix as created earlier for MapBiomas
mapbiomas_2015_reclass <- classify(mapbiomas_2015_cropped, reclass_m)

# use if you want to assign all 0 pixels to no data value:
#mapbiomas_reclass[mapbiomas_reclass == 0] <- NA

# save intermediate raster to save on computing time in future
writeRaster(mapbiomas_2015_reclass, "data/mapbiomas_brasil_2015_cropland.tif")
```

